Modeling costs and benefits of genomic wastewater surveillance

### 2023-05-12 paper/model-documentation.qmd
updating model documentation

# Background

@@

This paper develops a stochastic metapopulation disease transmission model and uses it to assess the conditions under which WBS systems can provide value to local public health decision makers who respond to a pandemic using surveillance data. We evaluate a range of scenarios that represent different potential pathogens and highlight scenarios under which WBS systems may or may not be cost-effective.

# Model

@@

```{mermaid}
%%| label: fig-model-diagram
%%| fig-cap: Model diagram
%%| fig-width: 6.5
%%| fig-align: center
flowchart LR
  S --> E --> P --> I --> R
  I --> l --> E
```

@@

## Nonpharmaceutical interventions {#npi}

@@

and the intervention level $\mathbf{l}_t$ is a jurisdiction-dependent nonpharmaceutical intervention level. $\mathbf{l}_t$ can be either specified as an exogenous variable or an endogenous variable controlled by local publich health departments.

## Controlling NPIs

In this model, local public health decision-makers follow epidemiological outcomes that are proportional to $\mathbf{I_t}$ (such as case counts) to decide the appropriate level of intervention. Thus, we specify a target NPI level $\mathbf{l}_t^*$ as a function of the disease state.

$$ 
\mathbf{l}_t^* = cI
$$ {#eq-targetnpi}

This formulation approximates phased mitigation plans where policymakers introduce or remove interventions depending on disease prevalence. While public health decision-makers strive to use up-to-date information to introduce and remove NPIs, they do not have perfect and instantaneous information about the disease state. In continuous time, we represent this process with a stock vector $\mathbf{L}$ representing the NPI level in effect

@@

$$
C = \int_{t_0}^{t_f}{f(\mathbf{l}_t\tau)}
$${#eq-cost}

where $f(x)$ is a function that maps a percent reduction in potentially infectious contacts to a monetary value. At the time of the intervention, decision-makers do not have precise information about the costs of interventions $\tau$ or its costs represented by function $f(x)$.

@@

-->
```
# Tables

>>>>

# Introduction

@@

This paper develops a stochastic metapopulation disease transmission model and uses it to assess the conditions under which WBS systems can provide value to local public health decision makers who respond to a pandemic using surveillance data. We evaluate a range of scenarios that represent different potential pathogens and highlight scenarios under which WBS systems may or may not be cost-effective.
<!--
Our results highlight that the value of WBS systems is contingent on a set of biological and societal parameters, including pathogen characteristics (i.e., transmissibility and severity), as well as the geographic scale of 
-->
# Methods

## Disease Transmission Model

@@

```{mermaid}
%%| label: fig-model-diagram
%%| fig-cap: Model diagram
%%| fig-width: 6.5
%%| fig-align: center
flowchart LR
  S --> E --> P --> I --> R
  I --> D
```

@@

### Nonpharmaceutical interventions {#npi}

@@

and the intervention level $\mathbf{l}_t$ is a jurisdiction-dependent nonpharmaceutical intervention level. $\mathbf{l}_t$ can be either specified as an exogenous variable or an endogenous variable controlled by local public health departments. Public health decision-makers follow epidemiological outcomes that are proportional to $\mathbf{I_t}$ (such as case counts) to decide the appropriate level of intervention. Thus, we specify a target NPI level $\mathbf{l}_t^*$ as a linear function of disease prevalence.

$$ 
\mathbf{l}_t^* = \min{(cI, l_{max})}
$$ {#eq-targetnpi}

This formulation approximates phased mitigation plans adopted during the COVID-19 pandemic, where policymakers introduce or remove interventions depending on disease prevalence or other severity indicators. While public health decision-makers strive to use up-to-date information to introduce and remove NPIs, they do not have perfect and instantaneous information on disease prevalence; hence actual policy behaviors lag epidemiological indicators. In continuous time, we represent this process with a stock vector $\mathbf{L}$ representing the NPI level in effect

@@

$$
C_{npi} = \int_{t_0}^{t_f}{f(\mathbf{l}_t\tau)}
$${#eq-cost}

where $f(x)$ is a function that maps a percent reduction in potentially infectious contacts to a monetary cost. At the time of the intervention, decision-makers do not have precise information on either the effects of interventions (expressed in our model as a percent reduction $100 * \tau$ in infectious contacts) or its costs represented by function $f(x)$. Even if public health decision makers could estimate the instantaneous costs of nonpharmaceutical interventions in real-time, they would not be able to observe the full costs of interventions that do not manifest during the time-frame of interventions (i.e., the life-time effect of school closures). Nevertheless, any attempt to value the benefits of surveillance systems require the specification of the costs of nonpharmaceutical interventions.

## Net monetary benefit (NMB)

The net monetary benefit of interventions can be expressed as

$$
E[NMB] = E[QALYs] * WTP - E[C_{npi}] - C_{surv} 
$${#eq-nbm}

where $QALYs$ represent the quality-adjusted life-years gained from non-pharmaceutical interventions, $WTP$ is the willingness-to-pay for one quality-adjusted life_year, $C_{npi}$ are the total costs of NPIs and $C_{surv}$ is the total cost of surveillance system employed.

# Valuing Wastewater surveillance information

In this analysis, WBS systems are not assumed to resolve uncertainty of an exogenous model parameter $\boldsymbol \theta$. If WBS systems did resolve those uncertainties, then it would be possible to estimate the value of information (VOI) provided by wastewater surveillance systems using an expected value of partial perfect information (EVPPI) approach. Rather, we assume that WBS systems change what signals are available to policymakers and hence *how* they respond to an endogenous model quantity. Hence, we define the value of wastewater surveillance information as simply the difference between the net monetary benefit of with and without surveillance

$$
V = E[NMB|WBS=1] - E[NMB|WBS=0]
$$ {#eq-wbsvalue}

@@

-->
```

# Results

What follows is an outline of the results section.

- Disease Elimination Scenarios

First, we demonstrate the value of genomic surveillance scenarios in conditions under which temporary disease elimination is possible. These scenarios represent disease dynamics observed in small geographic regions (cite https://www.nejm.org/doi/full/10.1056/NEJMc2025203 New Zealand).

{Figure with model dynamics with and without genomic surveillance system}

- Disease Control Scenarios

Early-warning systems can also be valuable in jurisdictions that do not achieve disease elimination.

# Appendix

## Tables
